---
title: "Simple Linear Regression"
author: "Jenocent Edwardraj"
date: "2024-09-22"
output:
  ioslides_presentation: default
---

<style type="text/css">
body p, div, h1, h2, h3, h4, h5 {
  color: black;
  font-family: Modern Computer Roman;
}
slides > slide.title-slide hgroup h1 {
  color: #8C1D40;
}
h2 {
  color: #8C1D40;
}
</style>

# 1: Title and Introduction

## (Part 1) - Simple Linear Regression

### What is Simple Linear Regression?


Simple Linear Regression is a statistical method used to understand and model the relationship between two variables: 

- **One independent variable** (predictor, denoted as \( x \)).

- **One dependent variable** (response, denoted as \( y \)).

It assumes that there is a linear relationship between the two variables, which can be represented by a straight line.

## (Part 2) - Why is it Important?

Linear regression is one of the most fundamental and widely used techniques in data analysis. It helps in:

- Predicting values of the dependent variable based on the independent variable.

- Understanding the strength and nature of the relationship between the two variables.

- Making future predictions, business decisions, or solving real-world problems.

## (Part 3) - Applications

- **Finance**: Predicting stock prices based on historical data.

- **Economics**: Forecasting economic indicators such as inflation or GDP.

- **Biology**: Modeling the relationship between a drug dosage and patient recovery rate.

- **Engineering**: Estimating material strength as a function of load applied.


# 2: Definition of Simple Linear Regression

## (Part 1) - How does Simple Linear Regression Work?

Simple linear regression models the relationship between two variables by fitting a linear equation to observed data. The model tries to minimize the difference between the observed values and the values predicted by the linear equation.

## (Part 2) - Equation
The equation for Simple Linear Regression is:

\[
y = \beta_0 + \beta_1 x + \epsilon
\]

Where:

- \( y \) is the **dependent variable** (what you are trying to predict).

- \( x \) is the **independent variable** (the predictor).

- \( \beta_0 \) is the **intercept** (the value of \( y \) when \( x = 0 \)).

- \( \beta_1 \) is the **slope** (the rate of change in \( y \) for each unit increase in \( x \)).

- \( \epsilon \) is the **error term** (captures the variability in \( y \) that cannot be explained by the model).

## (Part 3) - Key Concepts

- **Intercept (\( \beta_0 \))**: This is the point where the regression line crosses the y-axis. It represents the expected value of \( y \) when \( x \) is zero.
  
- **Slope (\( \beta_1 \))**: This represents the change in the dependent variable for each one-unit change in the independent variable. A positive slope means as \( x \) increases, \( y \) increases, and vice versa.
  
- **Residuals**: The difference between observed values and predicted values. Residuals help assess the accuracy of the model.


# 3: Assumptions of Simple Linear Regression
## Assumptions

1. **Linearity**: The relationship between the independent variable \(x\) and the dependent variable \(y\) is linear.

2. **Independence**: Observations are independent of each other.

3. **Homoscedasticity**: The variance of residuals (errors) is constant across all values of the independent variable.

4. **Normality**: Residuals are normally distributed.

5. **No Multicollinearity**: In simple linear regression, we only have one predictor, so this assumption is naturally met.

*Understanding these assumptions is crucial because violations can lead to biased or inefficient estimates.*

# 4: Example Data (ggplot2)

## (Part 1) - Scatterplot with Linear Fit of Generated Data

```{r, message=FALSE, echo = FALSE}
# Load necessary package
library(ggplot2)

# Generate random data
set.seed(123)
data <- data.frame(x = rnorm(100), y = rnorm(100))

# Scatterplot with linear fit
ggplot(data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Scatterplot with Linear Fit", x = "X", y = "Y") +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title
```

## (Part 2) - Description/Explanation
This scatterplot visualizes the relationship between the independent variable \( x \) and the dependent variable \( y \), with a fitted linear regression line. The model suggests how well the independent variable \( x \) predicts the dependent variable \( y \), capturing any linear trend in the data.


# 5: Interactive Plot (Plotly Plot)

## (Part 1) - Interactive Linear Regression Model with Plotly
```{r, message=FALSE, echo = FALSE}
# Load necessary package
library(plotly)

# Fit a linear model
model <- lm(y ~ x, data = data)

# Predict values based on the model
data$pred <- predict(model)

# Create interactive plot with plotly
p <- plot_ly(data, x = ~x, y = ~y, type = "scatter", mode = "markers") %>%
  add_trace(x = ~x, y = ~pred, mode = "lines", name = "Linear Fit")

# Display the plot
p
```

## (Part 2) - Description/Explanation
The interactive plot allows for dynamic exploration of the relationship between \( x \) and \( y \). You can hover over points to see the precise values and observe the linear trend, making it easier to understand the model's behavior and predictions.


# 6: Residuals vs Fitted plot

## (Part 1) - Validating Model Assumptions: Residuals vs Fitted Plot
```{r, message=FALSE, echo = FALSE}
ggplot(data, aes(x = model$fitted.values, y = model$residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title
```

## (Part 2) - Description/Explanation
This plot is a diagnostic tool to check the assumptions of the linear regression model. The residuals, which are the differences between observed and predicted \( y \) values, should be randomly scattered around zero. This indicates that the model fits the data well, with no obvious patterns in the residuals.


# 7: ggplot2 Diagnostic Plot (Residuals Histogram)

## (Part 1) - Histogram of Model Residuals for Diagnostic Evaluation
```{r, message=FALSE, echo = FALSE}
# Histogram of residuals
ggplot(data, aes(x = model$residuals)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  labs(title = "Residual Histogram", x = "Residuals", y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title
```

## (Part 2) - Description/Explanation
The histogram shows the distribution of residuals, which helps in assessing whether the residuals are normally distributed. A bell-shaped curve would suggest that the assumption of normality for residuals is met, an important aspect for accurate regression modeling.


# 8: Simple R Code for Linear Regression Summary

## (Part 1) - Summary of Linear Regression Model Output
```{r, message=FALSE, echo = FALSE}
# Show summary of the linear model
summary(model)
```

## (Part 2) - Description/Explanation
The linear model summary provides statistical insights into the relationship between \( x \) and \( y \). The coefficients, \( \beta_0 \) (intercept) and \( \beta_1 \) (slope), quantify the model's predictions, while the \( R^2 \) value measures the proportion of variance in \( y \) explained by \( x \). A high \( R^2 \) suggests a strong linear relationship between the two variables.


# Slide 9: Conclusion
## Summary of Simple Linear Regression
- Simple Linear Regression is a foundational technique to model relationships between two variables.

- The model assumes a linear relationship, normally distributed residuals, and constant variance.

- Evaluating residuals through diagnostic plots ensures that the model fits the data appropriately.

- Linear regression serves as a basis for more complex models and helps in making predictions or identifying trends.


